Highly Available Transactions: Virtues and Limitations

### 2013-03-31 writeup/conclusion.tex
PDB pass plus Alan comments

VIII. Conclusion (.5pg)

HATS represent a new class of database system:
* Can't get serializability, but can get other properties
* Key challenge: designing distributed algorithms and architectures for partial failure, asynchrony
* Rearchitecture of DBMSs and better understanding of applications are key

References (1pg)

>>>>

In this paper, we have explored the relationship between
multi-operation, multi-object transactions and high availability. We
have outlined requirements for high availability in a transactional
context, or Highly Available Transactions (HATs) and taxonomized
guarantees spanning both distributed systems and ACID database
literature. Many transactional guarantees are achievable with high
availability, while a few, like preventing Lost Update and Write Skew,
are not. To understand the benefits of high availability and low
latency, we classified the prevalence and cost of partitions and
network latency both in the literature and via experiment, analyzing
existing transactions to determine whether HATs were sufficient to
ensure application consistency, and validated intuitions about the
performance benefits of HAT on public cloud infrastructure.

Highly available operation represents a serious but often overlooked
consideration in distributed database designs. While the CAP Theorem
has had a large impact on highly available distributed data storage
design, our treatment of high availability and low latency in
transactional database systems has not kept pace. The key distributed
systems challenge of designing distributed algorithms and
architectures for partial failure and network asynchrony is largely
unaddressed in the literature on transaction processing systems. Fully
realizing the potential of HAT systems and bringing transactional
functionality to a large class of scalable data stores will require
further rearchitecture and study of existing database designs as well
as better understanding of application semantics and their
relationship to distributed consistency.

### 2013-03-31 writeup/evaluation.tex
PDB pass plus Alan comments

We now have an understanding of which guarantees are and are not
HAT-compliant. In this section, we begin to understand the impact of
this high availability on existing systems and applications as well as
briefly demonstrate HAT algorithms on public cloud
infrastructure. Specifically:

@@

\vspace{.5em}\noindent\textbf{Serializability} To establish a serial
order on transactions, algorithms for achieving serializability of
general-purpose read-write transactions in a distributed
setting~\cite{bernstein-book} may require at least one round trip time
(RTT) before committing. As an example, traditional two-phase locking
for a transaction of length $T$ requires $T$ \texttt{lock} operations
and at least one \texttt{unlock} operation.  In a distributed
environment, each of these operations requires coordination, either
with other database replicas or with a lock service. If the
coordination mechanism for locking is unavailable, transactions cannot
commit. Similarly, optimistic concurrency control requires
coordinating via a validation step, while deterministic transaction
scheduling~\cite{deterministic-scheduling} requires a
scheduler. Serializability under multi-version concurrency control
also requires checking for update conflicts.

In summary, perhaps unsurprisingly, the reliance on a globally agreed
total order necessitates a minimum of one round-trip to a designated
master or coordination service for each of these classic algorithms.
This is intuitively a consequence of serializability's lack of high
availability~\cite{davidson-survey} (or, as we have shown, reliance on
preventing Lost Update and Write Skew).  The cost of the (minimum one)
round trip will be determined by the deployment environment, as we saw
in Section~\ref{sec:motivation}.

\vspace{.5em}\noindent\textbf{Non-serializability} Many existing
distributed implementations of weak isolation are not highly
available. Lock-based proposals such as those used to provide weak
isolation in Gray's original proposal~\cite{gray-isolation} do not
degrade gracefully in the presence of partial failure. (Note, however,
that lock-based protocols \textit{do} offer the benefit of recency
guarantees.) While multi-versioned storage systems allow for a variety
of transactional guarantees, we have not seen proposals for truly weak
isolation (e.g., non-``tentative update'' schemes) in this
context. Chan and Gray's read-only transactions provide read-only
transactions with item-cut isolation, causal consistency, and
transactional atomicity (session PL-2L) but are unavailable in the
presence of coordinator failure~\cite{readonly}, similar to read-only
and write-only transactions more recently proposed by
Eiger~\cite{eiger}.  Bolt-on causal consistency~\cite{bolton},
Brantner's S3 database~\cite{kraska-s3}, and
Bayou~\cite{sessionguarantees} can all provide variants of session
PL-2L with high availability, but none provide this HAT functionality
without substantial modification; Swift~\cite{swift} is closest to
providing full HAT operation in a sticky context. Similarly, the
MDCC~\cite{mdcc} protocol offers Read Committed isolation with Lost
Update avoidance but is similarly unavailable due to its reliance on
preventing write conflicts.

As we have seen, it is possible to implement many guarantees weaker
than serializability---including guarantees that are achievable with
high availability---and still not achieve high availability.

@@

Thus far, we have largely ignored the question of which applications
require semantics that are unavailable to HATs. As we showed in
Section~\ref{sec:hats}, the main cost of choosing HATs comes in the
inability to prevent Lost Update, Write Skew, and provide recency
bounds. In this section, we attempt to understand when these
guarantees matter both abstractly and by the use of a representative
customer order transaction application patterned off of the TPC-C
benchmark~\cite{tpcc}.

\vspace{.5em}\noindent\textbf{Commutativity and Monotonicity} Recent
work on the CALM Theorem~\cite{calm}, Lattice-based state
update~\cite{blooml}, and Commutative and Replicated Data
Types~\cite{crdt} demonstrates that, if updates logically commute,
then they can often be safely performed in different orders at
different locations in a distributed system. Accordingly, as long as
all writes are delivered to all replicas, then a system executing
monotonic logic with commutative operators may not suffer from
application-level consistency anomalies as a result of Lost Update or
Write Skew anomalies. However, applications with non-monotonic state
mutation will not, in general, be able to maintain application-level
consistency constraints with HATs alone. Similarly, applications
requiring write visibility between clients should opt for
unavailability instead of possibly missing a write visibility
deadline.

\vspace{.5em}\noindent\textbf{TPC-C} Absent a necessary and sufficient
result describing exactly which application logics can be achieved
with high availability and to make these requirements more concrete,
we consider an example application: the TPC-C benchmark. Our goal is
\textit{not} to provide a performance analysis, as a HAT
configuration will be non-compliant. Rather, we examine the five
queries contained in the benchmarks---ostensibly representative of a
standard transaction processing application---to understand which are
achievable in a HAT context. Surprisingly, four of five can be
executed with HATs, while the fifth may require unavailablity.

TPC-C consists of five transactions, capturing operation of a
wholesale warehouse, including sales, payments, and deliveries. Two
transactions---\textit{Order-Status} and \textit{Stock-Level}---are
read-only and can be executed safely under HATs. Clients may read
stale data, but this does not violate TPC-C requirements and clients
will read their writes if they are sticky-available. Another
transaction type, \textit{Payment}, updates running balances for
warehouses, districts, and customer records as well as providing an
audit trail. The transaction is increment-only, so all balance
increase operations can commute, and TA allows the maintainence of any
foreign-key constraints (e.g., \texttt{UPDATE/DELETE CASCADE}).

While three out of five transactions have been easily achievable with
HATs, the remaining two transactions---\textit{New-Order} and
\textit{Delivery}---are not so simple. The New-Order transaction
places an order for a variable quantity of data items, updating
warehouse stock as needed. It selects a sales district, assigns the
order an ID number, adjusts the remaining warehouse stock, and writes
a placeholder entry for the pending order. The Delivery transaction
represents the fulfillment of a New-Order: it deletes the order from
the pending list, updates the customer's balance, updates the order's
carrier ID and delivery time, and updates the customer balance.

There are three challenges that only allow HATs to place non-compliant
New-Orders and to fail on Payments. First, New-Order requires a unique
ID number for each order. We can create a unique number by, say,
concatenating the client ID and a timestamp. However, the TPC-C
specification requires order numbers to be \textit{sequentially}
assigned, which requires preventing Lost Update. Accordingly, HATs
cannot provide compliant TPC-C execution but can maintain uniqueness
constraints. Second, the New-Order transaction decrements inventory
counts: what if the count becomes negative? Fortunately, TPC-C
New-Order restocks each item's inventory count (increments by 91) if
it would become negative as the result of placing an order. This means
that, even in the presence of concurrent New-Orders, an item's stock
will never fall below zero. This is TPC-C compliant, but a HAT system
might end up with substantially more stock than in a single-site
implementation. Third, delivery events are non-monotonic---deleting a
pending New-Order transaction---and should be idempotent in order to
avoid billing a customer twice; this implies a need to prevent Lost
Update. We can avoid this issue by moving the non-monotonicity to the
real world---the carrier that picks up the package for an order can
ensure that she is the only carrier who has done so---but cannot
provide a correct execution with HATs alone. However, according to
distributed transaction architects~\cite{helland}, these compensatory
actions are relatively common in real-world business processes.

Throughout execution, TPC-C also requires the maintainence of several
integrity constraints. For example, Consistency Condition 1 (3.3.2.1)
requires that each warehouse's sales count must reflect the sum of its
subordinate sales districts. This integrity constraint spans two
tables but, given the ability to update rows in both tables atomically
via transactional atomicity, can be easily maintained. Consistency
Conditions 4 through 12 (3.3.2.4-12) can similarly be satisfied by
applying updates atomically across tables. Consistency Conditions 2
and 3 (3.3.2.2-3) concern order ID assignment and are more
problematic, as we have discussed.

In summary, many---but not all---TPC-C transactions are well-served by
HATs. The two problematic transactions---New-Order and Payment---rely
on non-monotonic state update. The former transaction can be modified
to ensure ID uniqueness at the expense of sequential per-district ID
ordering, while the latter is an inherently non-monotonic action
requiring external compensation or stronger consistency protocols. We
expect that, especially for read-dominated workloads found in many
online services, HAT guarantees will suffice for many application
transactions. Finally, while TPC-C is not subject to multi-key
anomalies, we note that many TPC-E isolation tests, such as one that
requires simultaneously modifying a product description and its image,
are also passable using HATs.

>>>>

With an understanding of which guarantees are and are not
HAT-compliant, in this section, we analyze the implications of these
results for existing systems and applications and briefly study HAT
systems on public cloud infrastructure. Specifically:

@@

\vspace{.5em}\noindent\textbf{Serializability} To establish a serial
order on transactions, algorithms for achieving serializability of
general-purpose read-write transactions in a distributed
setting~\cite{bernstein-book} may require at least one round trip time
(RTT) before committing. As an example, traditional two-phase locking
for a transaction of length $T$ requires $T$ \texttt{lock} operations
and at least one \texttt{unlock} operation.  In a distributed
environment, each of these operations requires coordination, either
with other database replicas or with a lock service. If this
coordination mechanism is unavailable, transactions cannot safely
commit. Similarly, optimistic concurrency control requires
coordinating via a validation step, while deterministic transaction
scheduling~\cite{deterministic-scheduling} requires a
scheduler. Serializability under multi-version concurrency control
also requires checking for update conflicts. The reliance on a
globally agreed total order necessitates a minimum of one round-trip
to a designated master or coordination service for each of these
classic algorithms.  The cost of the (minimum one) round trip will be
determined by the deployment environment, as we saw in
Section~\ref{sec:motivation}; we will demonstrate this cost on public
cloud infrastructure in the next section.

\vspace{.5em}\noindent\textbf{Non-serializability} Many existing
distributed implementations of weak isolation are not highly
available. Lock-based proposals such as those used to provide weak
isolation in Gray's original proposal~\cite{gray-isolation} do not
degrade gracefully in the presence of partial failure. (Note, however,
that lock-based protocols \textit{do} offer the benefit of recency
guarantees.) While multi-versioned storage systems allow for a variety
of transactional guarantees, we have not seen proposals for truly weak
isolation (e.g., non-``tentative update'' schemes) in this context.
The MDCC~\cite{mdcc} protocol offers Read Committed isolation with
Lost Update avoidance but is similarly unavailable due to its reliance
on preventing write conflicts. Chan and Gray's read-only transactions
provide read-only transactions with item-cut isolation, causal
consistency, and transactional atomicity (session PL-2L~\cite{adya})
but are unavailable in the presence of coordinator
failure~\cite{readonly}, similar to read-only and write-only
transactions more recently proposed by Eiger~\cite{eiger}; Causal
Serializability offers a similar model (with unavailable
implementation): causal consistency with a variant of Read Uncommitted
between transactions that write to the same data
item~\cite{raynal-causal}.  Bolt-on causal consistency~\cite{bolton},
Brantner's S3 database~\cite{kraska-s3}, and
Bayou~\cite{sessionguarantees} can all provide variants of session
PL-2L with high availability, but none provide this HAT functionality
without substantial modification; Swift~\cite{swift} is closest to
providing full HAT operation in a sticky context. As we have seen, it
is possible to implement many guarantees weaker than
serializability---including guarantees that are achievable with high
availability---and still not achieve high availability.

@@

Thus far, we have largely ignored the question of which applications
require semantics that are unavailable to HATs. As we showed in
Section~\ref{sec:hats}, the main cost of choosing HATs comes in the
inability to prevent Lost Update, Write Skew, and provide recency
bounds. In this section, we attempt to understand when these
guarantees matter both abstractly and in a representative
transactional application based on the TPC-C benchmark~\cite{tpcc}.

\vspace{.5em}\noindent\textbf{Commutativity and Monotonicity} Recent
work on the CALM Theorem~\cite{calm}, Lattice-based state
update~\cite{blooml}, and Commutative and Replicated Data
Types~\cite{crdt} demonstrates that, if updates logically commute,
then they can often be safely performed in different orders at
different replicas. Accordingly, as long as all writes are delivered
to all replicas, then a system executing monotonic logic with
commutative operators may not suffer from application-level
consistency anomalies as a result of Lost Update or Write Skew
anomalies. However, applications with non-monotonic state mutation
will not, in general, be able to maintain application-level
consistency constraints with HATs alone. Moreover, applications with
requiring bounded update visibility latency should opt for
unavailability.

\vspace{.5em}\noindent\textbf{TPC-C} To better understand the impact
of HAT-compliance in an application context, we consider a concrete
application: the TPC-C benchmark. Surprisingly, four of five
transactions can be executed with HATs, while the fifth may require
unavailablity.

TPC-C consists of five transactions, capturing the operation of a
wholesale warehouse, including sales, payments, and deliveries. Two
transactions---\textit{Order-Status} and \textit{Stock-Level}---are
read-only and can be executed safely under HATs. Clients may read
stale data, but this does not violate TPC-C requirements and clients
will read their writes if they are sticky-available. Another
transaction type, \textit{Payment}, updates running balances for
warehouses, districts, and customer records as well as providing an
audit trail. The transaction is increment- and append-only, so all
balance increase operations commute, and TA allows the
maintainence of any foreign-key constraints (e.g.,
\texttt{UPDATE/DELETE CASCADE}).

While three out of five transactions are easily achievable with
HATs, the remaining two transactions---\textit{New-Order} and
\textit{Delivery}---are not as simple. The New-Order transaction
places an order for a variable quantity of data items, updating
warehouse stock as needed. It selects a sales district, assigns the
order an ID number, adjusts the remaining warehouse stock, and writes
a placeholder entry for the pending order. The Delivery transaction
represents the fulfillment of a New-Order: it deletes the order from
the pending list, updates the customer's balance, updates the order's
carrier ID and delivery time, and updates the customer balance.

New-Order presents two challenges: ID assignment and stock
maintainence. First, each New-Order transaction requires a unique ID
number for the order. We can create a unique number by, say,
concatenating the client ID and a timestamp. However, the TPC-C
specification requires order numbers to be \textit{sequentially}
assigned within a district, which requires preventing Lost
Update. Accordingly, HATs cannot provide compliant TPC-C execution but
can maintain uniqueness constraints. Second, the New-Order transaction
decrements inventory counts: what if the count becomes negative?
Fortunately, TPC-C New-Order restocks each item's inventory count
(increments by 91) if it would become negative as the result of
placing an order. This means that, even in the presence of concurrent
New-Orders, an item's stock will never fall below zero. This is TPC-C
compliant, but a HAT system might end up with substantially more stock
than in a non-HAT-compliant implementation.

Delivery is challenging due to non-monotonicity. Each Delivery deletes
a pending order from the NewOrder Table and should be idempotent in
order to avoid billing a customer twice; this implies a need to
prevent Lost Update. We can avoid this issue by moving the
non-monotonicity to the real world---the carrier that picks up the
package for an order can ensure that she is the only carrier who has
done so---but cannot provide a correct execution with HATs
alone. However, according to distributed transaction
architects~\cite{entitygroup}, these compensatory actions are
relatively common in real-world business processes.

Throughout execution, TPC-C also requires the maintainence of several
integrity constraints. For example, Consistency Condition 1 (3.3.2.1)
requires that each warehouse's sales count must reflect the sum of its
subordinate sales districts. This integrity constraint spans two
tables but, given the ability to update rows in both tables atomically
via transactional atomicity, can be easily maintained. Consistency
Conditions 4 through 12 (3.3.2.4-12) can similarly be satisfied by
applying updates atomically across tables. Consistency Conditions 2
and 3 (3.3.2.2-3) concern order ID assignment and are more
problematic, as we have discussed.  Finally, while TPC-C is not
subject to multi-key anomalies, we note that many TPC-E isolation
tests, such as one that requires simultaneously modifying a product
description and its image, are also passable using HATs.

In summary, many---but not all---TPC-C transactions are well-served by
HATs. The two problematic transactions---New-Order and Payment---rely
on non-monotonic state update. The former transaction can be modified
to ensure ID uniqueness at the expense of sequential per-district ID
ordering, while the latter is an inherently non-monotonic action
requiring external compensation or stronger consistency protocols. We
expect that, especially for read-dominated workloads found in many
online services, HAT guarantees will suffice for many application
transactions.

### 2013-03-31 writeup/futurework.tex
PDB pass plus Alan comments

We believe that there are several promising avenues for future work in
Highly Available Transaction systems:

\textbf{What are the limits of HAT systems, and how are they defined?}
In this work, we have taxonomized several well-defined and
well-documented consistency properties. This is useful as it
contextualizes decades of existing database and distributed systems
design and algorithms but does not provide guidance as to what
consistency models are the ``strongest''---or admit the smallest set
of executions---within each availability class. A recent report from
UT Austin shows that no model stronger than causal consistency can be
achieved in a sticky highly available, one-way convergent
system~\cite{cac}, but there are infinite highly available and sticky
highly available models to consider. For example, always returning the
value $1$ in response to reads is incomparable with (i.e., neither
stronger nor weaker than) causal consistency: returning $1$ does not
respect causality, and causal consistency returns values other than
$1$. Always returning the value $2$ is similarly incomparable, and, as
an example, we can enumerate models that return all integers and real
numbers--an uncountably infinite set of models. Given these infinite
models, are there meaningful properties that delineate the boundaries
between classes of highly available systems? Is there a
``meta-property'' that exactly characterizes what properties are
achievable with high availability or sticky high availability? This is
an area of ongoing research within the Berkeley HAT project.

@@

\textbf{Where do semantics-based concurrency control and weakened
  failure models lie in the HAT landscape?} The vast amount of
literature on semantics-based concurrency control provides hints as to
when coordination can be avoided while providing ``strongly
consistent'' outcomes for limited failure scenarios. We have performed
a narrow case study in this work, but a general taxonomization of
techniques such as escrow transactions~\cite{escrow},
Sagas~\cite{sagas}, and other long-running transaction
support~\cite{gray-virtues} could help answer the challenge of
programming in a weakly consistent environment. Additionally, HAT
guarantees can be strengthened by considering failure models with,
say, bounded asynchrony or service-level agreements from underlying
hardware. Additionally, while we have focused on systems that provide
a single availability mode, there are several interesting hybrids to
consider. For example, a system may provide ``best-effort'' stickiness
(or serializability!) in the absence of partitions, but, in the
presence of partitions, ``lose'' guarantees by having clients
reconnect to an available replica. This strategy provides stronger
guarantees in the absence of partitions but, to provide availability,
sacrifices them in the event of partitions.

>>>>

There are several promising avenues for future work in HAT systems:

\textbf{What are the limits of achievable HAT models, and how are they
  defined?}  In this work, we have taxonomized several well-defined
and well-documented semantic properties. This is useful as it
contextualizes decades of existing database and distributed systems
design and algorithms but does not provide guidance as to what
consistency models are the ``strongest''---or admit the smallest set
of executions---within each availability class. A recent report from
UT Austin shows that no model stronger than causal consistency can be
achieved in a sticky highly available, one-way convergent
system~\cite{cac}, but there are infinite highly available and sticky
highly available models to consider. For example, always returning the
value $1$ in response to reads is incomparable with (i.e., neither
stronger nor weaker than) causal consistency: returning $1$ does not
respect causality, and causal consistency returns values other than
$1$. Always returning the value $2$ is similarly incomparable, and, as
an example, we can enumerate models that return all integers and real
numbers--an uncountably infinite set of models. Given these infinite
models, are there meaningful properties that delineate the boundaries
between classes of highly available systems? Is there a
``meta-property'' that exactly characterizes what properties are
achievable with high availability or sticky high availability? This is
an area of ongoing research within the Berkeley HAT project.

@@

\textbf{Where do semantics-based concurrency control and weakened
  failure models lie in the HAT landscape?} The vast amount of
literature on semantics-based concurrency control provides hints as to
when coordination can be avoided while providing ``strongly
consistent'' outcomes for limited failure scenarios. We have performed
a narrow case study in this work, but a general taxonomization of
techniques such as escrow transactions~\cite{escrow},
Sagas~\cite{sagas}, and other long-running transaction
support~\cite{gray-virtues} could help answer the challenge of
programming in a weakly consistent environment. HAT guarantees can
also be strengthened by considering weaker failure models such as
bounded asynchrony or models in which the underlying hardware provides
explicit service-level agreements. Additionally, while we have focused
on systems that provide a single availability mode, there are several
interesting hybrids to consider. For example, a system may provide
``best-effort'' stickiness (or serializability) in the absence of
partitions, but, in the presence of partitions, ``lose'' guarantees by
having clients reconnect to an available replica. This strategy
provides stronger guarantees in the absence of partitions but, to
provide availability, sacrifices them in the event of partitions.
